Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-12-01 response_to_reviewers.txt

Reviewer 2

Page 1

27: The spelling 'multimodal' seems to be more common than 'multi-modal' (and would be in line with other uses in your paper; e.g. page 3, line 12).
* Corrected.

30: The word 'the' seems to be missing before 'study'.
* Corrected.

32: The word 'the' seems to be missing before 'Ricci-Ollivier'.
* Corrected.

36: Here, I would suggest clarifying the phrase 'and n their number of leaves' a bit. In its current state, it might not be clear from the wording whether each tree has n leaves or whether n is the total number of leaves of all trees.
* Revised to "... and n is the number of leaves in each tree, ...".

44-46: The phrasing 'We show ... showing ...' sounds a bit odd because 'show' is used twice.
* Revised "showing" to "demonstrating".

Page 2
10: It looks like some kind of connecting phrase is missing between 'particular' and 'properties' (e.g.  'with respect to').
* We added the suggested connecting phrase.

11: The word 'the' seems to be missing before 'investigation'.
* Corrected.

52-53: As you start this part of this sentence with 'both', an 'and' seems to be more fitting in 'leaves [12, 13] or for SPR'.
* Corrected.

54: Do you rather mean 'an infinite number of leaves' or 'infinitely many leaves' instead of 'infinite leaves'?
* Corrected to "infinitely many leaves"

@@

23-25: The case distinction requires that the extra root node ρ has depth -1, the actual root depth 0 and so on, but this is stated nowhere in the paper.
* We added a sentence defining the depth of the rho node to be -1 to the
	preliminaries: "The depth of \rho is considered to be -1, the depth of
	the original root is 0, and the depth of each other node is given by
	its distance from the original root.

33: Probably, the set-difference operator ('\') is supposed to denote the pruning (including the merging of the incoming and outgoing edge of the resulting unary node). Unfortunately, this is not explained. A short explanation would make this more obvious to the reader.
* We have added an explanation here: "This implies that both T \U = S \U
	and T \V = S \V by the definition of an rSPR operation, where \ here
	indicates removing the subtree and its parent edge and then replacing
	its former parent node and both adjacent edges with a single edge."

Page 12

15: It looks like that '-th' is missing behind the '|r|' in order to turn it into an ordinal number.
* Corrected.

23: What is the unit of the 'O(n) space'?
* Bits of memory, as is standard for the Big-O notation with respect to space.

@@

39:
1. There seems to be an inconsistency between the definition of internal edges and their number stated in this line. On page 5, internal edges are defined as those that do not contact a leaf or ρ. Then, a 'rooted tree with k leaves' cannot have 2(k-1) internal edges as stated here (because this is the number of all edges (also those leading to leaves), disregarding the extra root edge). So, either the definition of 'internal edge' has to be adopted or the word 'internal' has to be dropped here.
* We have dropped the word "internal" here.

2. The white space after 'edges' is missing.
* Corrected.

Page 20

9-19: While Corollary 5.4 appears correct, the given justification seems to be a bit problematic.  Based on the definition of γ and the excluded (subtracted) moves, it appears that these γ moves still include some which move R. For example, from the set of all possible moves for T (deg(T)), only some moves into R (2kb, case (ii)) and some moves of R's parent subtree (2(j-1), case (iv)) are excluded.  However, it is then stated that these moves 'can be applied to T and S with the same source and target nodes'. The minor (but arguable) problem concerns the part 'same source nodes': When are they the same? Is the parent node of R the same although its position differs in T and S? If not, then not all γ moves can be applied with the same source nodes. The bigger problem is created by the fact that moves moving R with the same target node eliminate the only difference between T and S and, thus, the difference between the obtained neighbours is zero (not one).  Overall, it seems to be a bit more complicated to describe how the desired (T',S') pairs can be obtained.
* Revised the first part with "We consider the parent node of R in both T and S to be the same source node for this purpose.". 
* TODO: We can either change the distance bound here to <= 0 or subtract
	out the distance 0 pairs. Check which is better after completing the
	other revisions. The argument I added after Corollary 6.6 that the Lemma 6.5 pairs of trees have negative curvature needs to be checked. Lemma 6.5 might need to take a change here in to account.

14: The phrase 'the pairings are disjoint' might need some more explanation as 'disjoint' might get interpreted in several ways (e.g. no tree is allowed to occur in more than one pairing vs. the pairings are pairwise distinct).
* Revised to note "...  each neighbor of T is paired
with a single neighbor of S, and vice versa. Note that a tree may be a
neighbor of both T and S and thus involved in two different pairings."

37: The word 'and' seems to be missing before 'then' .
* Corrected.

Lemma 5.5:
1. Concerning the proof of (i): it might be helpful to elaborate on why a case distinction is
necessary (Why is it necessary to increase j? Why are the subcases of (not) moving the additional leaf necessary?) in order to clarify the line of argument.
* We have revised the sentence introducing the cases to state "There are two cases, depending on whether the subtree of k leaves is moved to the root (thereby introducing a new level to the tree) or not.

@@

51-53: Using U as a subtree while it also denotes a whole tree in the neighbourhood of T and S just a few lines before can be slightly confusing.
* Indeed, thanks. We changed the second use of U to V and u' to v'.

Page 22

30-32: The statement on the proportionality of small moves and small subtrees seems to be a bit troubled.

1. For example, if both children of the root are not small, each small subtree with one leaf contributes 2n-6 moves, while each small subtree with two leaves contributes 2n-8 moves.
So, what is the factor of proportionality?
* Proportional was too strong a word here, so we changed "directly proportional to" to "determined by". We also added the sentence "Moreover, every tree has n 1-leaf subtrees." to make the line of reasoning more clear.

2. In the phrase 'where one of the root's children is small', is the intended meaning exactly one or at least one? Assuming, e.g., 2n-6 as the factor of proportionality (coming from (the majority of) the leaves), the statement does not hold in general (neither for exactly nor for at least). Consider, e.g., a tree with a leaf as the only small child of the root and at least three small subtrees with two leaves (for the case exactly) respectively the balanced tree with four leaves (for the case at least).
* See the previous response.

@@

36: Writing 'one 2-leaf subtree' instead of '1 2-leaf subtree' appears more readable (and is more consistent with the previous use of numerals).
* We have made the suggested change.

38-42:

1. Since Lemma 5.7 shall hold for n ≥ 3, the stated number of small moves (2n^2-4n-6) is not correct for all intended n. The reason is that for the ladder tree with three leaves, the small subtree with two leaves is also a child of the root and, thus, contributes 2n-3-3 (not 2n-3-5) moves. Hence, a case distinction seems to be necessary here and for the inequality in the subsequent paragraph (lines 40-42)
* To keep the lemma and proof simple (and because the n=3 tree space is trivial) we changed the assumption to n > 3, matching the subsequent lemmas.

2. The chain of (in)equalities in lines 40-42 does not run through, because adding (24-8n) leads to the n-free term '+12' (not '+16'). But this can be 'repaired' through the necessary case distinction n = 3 vs. n ≥ 4).
* See our response above.

45+55: Stating the explanation of the size '(in terms of the number of leaves)' directly in line 45 possibly avoids some confusion (which can occur when reading Lemma 5.8 while having the typical notion of size in mind).
* We added the statement to line 45.

54-55: The statement 'a ladder tree has the greatest average moved subtree size' does not seem to hold for all n ≥ 4. For n = 4, the ladder tree has an average moved subtree size of 1, while it is 4/3 for the balanced tree. Similarly, the balanced tree has a larger one for n = 8 (1.55 vs. 1.39).  This might pose a problem to the proof because it based on the idea that the ladder tree leads to themaximum.
* There appears to be some confusion on what constitutes a moved subtree here.  From our distribution of rSPR moves, the balanced tree with n = 4 leaves does not move any subtrees of size 2. There are moves of such trees that are equivalent to moves of leaves which appears to be the source of the confusion. As such, the average moved subtree size for n = 8 is 1.4 for the ladder tree and approximately 1.379 (160/116) for the balanced tree. We have added the following explanation to make this more clear to the reader:

"Now, we bound the average size of the moved subtrees (in terms of the number of leaves) from rSPR moves according to our distribution. Note that this average ignores some possible moves that are equivalent to moves of smaller subtrees. For example, from the balanced tree with 4 leaves, we do not consider 22moves of the root’s 2-leaf child subtrees because they are equivalent to moves of the leaves, as explained by our distribution."

@@

Page 23

11-13: The vertical bars denoting the size of N i (T) are missing at all three occurrences.
* Corrected.

11-20:

1. It is stated that |N_i(T)| ≤ 2n-3-5 for i ≤ 2, but in a ladder tree is, e.g., |N_1(T)| = (n-1)(2n-1-5) + (2n-1-3) > 2n-3-5.
2. Also, it is stated that |N_i(T)| > 2n-3-5 for i > 2, but in a ladder tree is, e.g., |N i (T)| = 1 * 2n- (2i-1)-5 < 2n-3-5 for 2 ≤ i ≤ n-2.
3. Therefore, also the statements on decreasing / increasing the actual values to the bound 2n-3-5 are the wrong way round. Possibly, this can be fixed by swapping the inequality signs.
* Corrected.

4. Is it actually obvious enough that setting all |N_i (T)| to 2n-3-5 leads to an upper for the quotient representing the average moved subtree size? A short explanation might make this point clearer.
* We have moved the reference to Lemma 5.7 earler (upon which this argument relies) and revised the proof for clarity.

5. In the last sentence of the proof, the weighted average uses a proportion of 1⁄2 for both small and large subtree (which is okay), but why is the average moved subtree size of small substrees assumed to be one? Shouldn't the one small subtree with two leaves (which always exists) make it slightly larger than one?
* Yes, we are reducing the contribution of the small subtrees to achieve an upper bound.

@@

19: The innermost pair of parentheses around 'O(n)' is not necessary.
* Corrected.

30: Replacing the word 'a' before 'distance' by 'at' or 'of' conveys the meaning better.
* Replaced with "of".

31: The equality sign in the equation might be less fitting than a <, because it is previously (line 25) stated that S_n has a strictly larger neighbourhood.
* It is counterintuitive at first glance, but the equal sign is correct
	because T_n and S_n differ by an O(kn) factor.

46-47: Shortly explaining how the bounds for the distance between neighbours of T and S lead to bounds for W_1 (m_T, m_S ) could make it easier for the reader to grasp the idea how the proof works.
* We have expanded the proof with such an explanation.

Page 25
29: Possibly it should be 'each other' instead of 'each either'.
* Corrected.

40-41: The actual meaning of the phrase 'a greater understanding of the distribution of tree neighborhoods with respect to one another' is not very clear (especially due to the underlined part).
* Revised to "However, we suspect that this will require a greater understanding of the distribution of pairwise distances between tree neighborhoods than is currently known.".

@@

Page 26

10: A short remark on how the term '1 + μ' is actually derived would make this passage easier / faster to comprehend
* Revised to "Pairs of neighbors of adjacent trees are at most distance 3 apart, so μ of the mass is moved a distance at most 3. The remaining 1−μ of the mass is moved between adjacent trees. Thus, we can bound the mass transport cost from above by 3μ + 1(1 − μ) = 1 + 2μ."

@@

14: Following Corollary 5.4 it should be 'at least γ' and not 'at most γ' paired trees (and, otherwise, the argumentation is not sensible at this point).
* Corrected.

19: The rationale for bounding the degree by using the maximum possible absolute difference (irrespective of the actual move) seems a bit unclear. Currently, it does not look like the line of argument leads to the maximum μ: using the maximum degree of T (instead of the actual, potentially smaller, value) does not result in the maximum μ because it leads to an unnecessarily large denominator.
* This pair of trees has the maximum possible degree change and S has the smallest possible degree. Any pair of trees with a smaller denominator will necessarily have a numerator that is reduced by at least the same amount, resulting in a smaller value of mu. We've added a sentence to this effect to make this more clear.

@@

2. It looks like the factors 'k' and 'a' are swapped after inserting their values. While this would not be wrong, it is a potential source of confusion.
* Corrected

39-40: Providing a proof or at least a clear rationale for the statement 'the pair of trees used in the proof of Lemma 6.5 will always have negative curvature, for all n ≥ 7' would lend more weight to it.
* We added such a statement

Page 27

35: The upper bound does not seem to follow as analogously to the lower bound as it is stated.  Using the inequality stated in line 32 and proceeding analogously to the lower bound leads to κ p (T , S ) 2 ⩽ ...⩽ κ (T , S)+ p p∗max (∣ N (T )∣ , ∣ N (S )∣) where the factor p in the denominator causes a problem, because simply dropping it (in order to arrive at the bound from the lemma) bounds the value in the wrong direction (as it is at most 1).
* Indeed, the stated observation was incorrect by a p factor in the numerator, because a fraction of p (the amount of mass that is moved) divided by the number of neighbors is the amount of mass that is moved from e.g. T to S and can therefore be matched with the lazy portion that remains at s. We have corrected the error.

@@

26: Within the min-term of the last case, w is the argument of all four probability densities.  Probably, two of them should have z as their argument.
* Corrected.

29-33: Using W 1 (resp. W' 1 ) without the arguments for the values of the Wasserstein distance is slightly imprecise.
* Corrected.

50: The word 'the' seems to be missing before 'understanding'.
* Corrected.

55: The statement 'rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth' seems a bit mixed up, because (according to the proven lemma) the changes depend linearly on the product (which in turn is quadratic in the number of leaves).
* Revised to "Moreover, we proved that rSPR graph degree changes depend on the product of the size of the regrafted subtree with its change in depth. This product is quadratic in the number of leaves in the trees, the same order as the degree."

### 2016-12-01 curvature.tex
most comments addressed, some figure comments remaining, two other TODOs remaining

	There are two cases, depending on whether the subtree of $k$ leaves is moved to the root or not.
    If not, then we set $j=1$ and split the remaining $n-b-i-j = n-2$ leaves between $k$ and $a$ in as balanced a way as possible, giving (i).
	Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the root-most leaf of the tree.

@@

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 3$ leaves.
	Let $N_S(T)$ be the neighbors of $T$ induced by small rSPR moves.
	Then $\frac{\size{N_S(T)}}{\size{N(T)}} > \frac{1}{2}$.
	\label{lem:spr_size}
\end{lem}
\begin{proof}
	Recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	Each subtree at a depth $ > 1$ from the root with $x$ nodes induces $2n - x - 5$ rSPR neighbors, while the children of the root induce $2n - x - 3$ neighbors.
	This implies that the number of small rSPR moves is directly proportional to the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.

@@

Now, we bound the average size of the moved subtrees (in terms of the number of leaves).

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 4$ leaves.
	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 3}{4} .$$
	\label{lem:spr_size_mean}
\end{lem}
\begin{proof}
	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
	There are $n$ subtrees with one leaf, and one subtree each with $2,3, \ldots n - 1$ leaves.
	We ignore the root subtree, which does not induce any rSPR moves.
	Thinking of the ratio as a weighted average of $i$, we can split the numerator and the denominator of the average into terms with $i \leq 2$ and those with $i > 2$.
    $\size{N_i(T)} \leq 2n - 3 - 5$ for the first set of terms, and $\size{N_i(T)} > 2n - 3 - 5$ for the second, so by replacing $\size{N_i(T)}$ in each case by $2n-3-5$ we decrease the weight for $i \leq 2$ and increase the weight on $i > 2$.
	Thus, by doing this replacement we obtain a upper bound on the mean moved subtree size.
	This bound, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a small subtree.
	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.

@@

\begin{lem}
	\label{lem:curvature_distance_bound}
	Let $T$ and $S$ be two trees. Then:
	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
\end{lem}
\begin{proof}
	Observe that the distance between neighbors of $T$ and $S$ is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
	For the curvature upper bound, we then have $\curvature{T,S} \le 1 - \frac{\dspr{T,S} - 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
	The lower bound follows similarly.
\end{proof}

@@

It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees, because neighborhood effects decrease with respect to the increasing distance.
However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
Moreover, nearly all of the neighbors of these pairs approach each other.
We thus conjecture the following:
\begin{conj}
	Let $k_n$ be the maximum curvature between two trees with $n$-leaves.
	Then:
	\begin{enumerate}
		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
		\item $ k_n \sim \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
	\end{enumerate}
\end{conj}
Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
However, we suspect that this will require a greater understanding of the distribution of tree neighborhoods with respect to one another than is currently known.

@@

\begin{lem}
\label{lem:curvature_lower_bound}
	The curvature between adjacent trees with $n$ leaves is at least
	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
\end{lem}
\begin{proof}
    In light of Corollary \ref{cor:paired_neighbors}, the optimal mass transport cost is maximized (and therefore curvature minimized) across adjacent trees $T$ and $S$ by a combination of two effects: trees that cannot be paired at distance $1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
		As we will show, these effects can be maximized simultaneously.
    To bound these effects, let $\mu$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
    We can bound the mass transport cost from above by $1 + 2\mu$ because pairs of neighbors of adjacent trees are at most distance 3 apart.
	This gives a lower bound of $1 - (1 + 2\mu) / 1) = -2\mu$ on the curvature.

	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
	By Corollary~\ref{cor:paired_neighbors}, there are at least $\overlap := \degree{T} -2ka - 2(i-1)$ paired trees, bounding the former effect.
	We now construct a pair of trees that maximizes both effects.
	Let $S$ be the ladder tree with degree $3n^2 - 13n + 14$ and $T$ be the adjacent tree constructed by moving the lower $\floor{\frac{n}{2}}$ leaves of $S$ to the root.
	$T$ has degree at most $3.5n^2 -15n + 16$.
	There are thus $2ka + 2(i-1) = 2\left(\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} +(1-1)\right) \le \frac{1}{2}n^2 - n$ unpaired neighbors, the maximum possible.
	Moreover, as shown by Lemma~\ref{lem:degree_change} this pair of trees obtains the maximum (absolute and relative) degree change.
	Thus, the maximum $\mu$ is:
	$$\frac{\frac{1}{2}n^2 - n}{3.5n^2 - 15n + 16}.$$
	The claim follows from multiplying this value by $-2$.
\end{proof}

@@

Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
We also note that the lower bounds in this section do not follow from the more general setting described in \citep{Jost2013-ce}.
However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.

@@

	For the upper bound, we observe that $W_1'(T,S) \ge$
	\begin{align*}
		p &W_1(T,S) + (1-p) \, \dspr{T,S} - \frac{2}{\max(\degree{T}, \degree{S})},
	\end{align*}
	as at most $1 / \max(\degree{T},\degree{S})$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
	The upper bound then follows analogously to the lower bound.
	Moreover, no mass can remain at $T$ or $S$ when $\dspr{T,S} > 1$, in which case the curvatures are equal.
\end{proof}

@@

\section{Conclusion and future work}
In summary, we have gone beyond graph diameter and vertex degree to substantially advance the understanding of the phylogenetic rSPR graph.
We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth.
As a result of this dependence, some areas of the rSPR graph are locally negatively curved but the graph tends toward flatness with respect to the majority of rSPR moves which move small subtrees.
Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.

>>>>

	There are two cases, depending on whether the subtree of $k$ leaves is moved to the root (thereby introducing a new level to the tree) or not.
    If not, then we set $j=1$ and split the remaining $n-b-i-j = n-2$ leaves between $k$ and $a$ in as balanced a way as possible, giving (i).
	Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the root-most leaf of the tree.

@@

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 4$ leaves.
	Let $N_S(T)$ be the neighbors of $T$ induced by small rSPR moves.
	Then $\frac{\size{N_S(T)}}{\size{N(T)}} > \frac{1}{2}$.
	\label{lem:spr_size}
\end{lem}
\begin{proof}
	Recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	Each subtree at a depth $ > 1$ from the root with $x$ nodes induces $2n - x - 5$ rSPR neighbors, while the children of the root induce $2n - x - 3$ neighbors.
	Moreover, every tree has $n$ 1-leaf subtrees.
	This implies that the number of small rSPR moves is determined by the number of small subtrees, with a possible difference of two moves for a tree where one of the root's children is small.

@@

Now, we bound the average size of the moved subtrees (in terms of the number of leaves) from rSPR moves according to our distribution. Note that this average ignores some possible moves that are equivalent to moves of smaller subtrees. For example, from the balanced tree with 4 leaves, we do not consider moves of the root's 2-leaf child subtrees because they are equivalent to moves of the leaves, as explained by our distribution.

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 4$ leaves.
	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 3}{4} .$$
	\label{lem:spr_size_mean}
\end{lem}
\begin{proof}
	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
	There are $n$ subtrees with one leaf, and one subtree each with $2,3, \ldots n - 1$ leaves.
	We ignore the root subtree, which does not induce any rSPR moves.
	Thinking of the ratio as a weighted average of $i$, we can split the numerator and the denominator of the average into terms for each subtree contributing to the values for $i \leq 2$ and those with $i > 2$.
    The contribution to $\size{N_i(T)}$ is $\geq 2n - 3 - 5$ for the first set of terms, and the contribution to $\size{N_i(T)} < 2n - 3 - 5$ for the second.
		By replacing the contribution to $\size{N_i(T)}$ in each case by $2n-3-5$ we decrease the weight for $i \leq 2$ and increase the weight on $i > 2$.
		Note that this replacement does not modify the $i = 2$ term (which was already exactly $2n - 3 - 5$ and decreases the weight of only the smallest terms (which must already be below the average size of the moved subtrees), so this provides an upper bound on the mean moved subtree size.
	The bound for large subtrees only is $(3 + (n-2)) / 2 = (n + 1) / 2$.
	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a small subtree.
	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.

@@

\begin{lem}
	\label{lem:curvature_distance_bound}
	Let $T$ and $S$ be two trees. Then:
	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
\end{lem}
\begin{proof}
	Observe that the distance between neighbors of $T$ and $S$ is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
	Then, any mass transported between a probability distribution $m_T$ on neighbors of $T$ and $m_S$ on neighbors of $S$ is transported a distance at least $\dspr{T,S} - 2$.
	This implies that the mass transport cost $W_1(m_T,m_S)$ is at least $\dspr{T,S} - 2$, for any $m_T$ and $m_S$.
	For the curvature upper bound, we then have $\curvature{T,S} \le 1 - \frac{\dspr{T,S} - 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
	The lower bound follows similarly by using the maximum distance of $\dspr{T,S} + 2$ .
\end{proof}

@@

It is more difficult to obtain a closer bound on the maximum curvature of nonadjacent trees.
Lemma~\ref{lem:curvature_distance_bound} suggests that more distant pairs of trees should have smaller curvatures than close trees, because neighborhood effects decrease with respect to the increasing distance.
However, our experiments with $n \le 7$ suggest that maximum curvature tends to increase with distance (with respect to a fixed $n$), as a far greater fraction of the neighbors approach each other as the distance increases.
Indeed, for $5 \le n \le 7$ the maximum curvature is obtained by pairs of trees at one less than the maximum distance.
Moreover, nearly all of the neighbors of these pairs approach each other.
We thus conjecture the following:
\begin{conj}
	Let $k_n$ be the maximum curvature between two trees with $n$-leaves.
	Then:
	\begin{enumerate}
		\item $k_n \le \frac{2}{\Delta_{\text{rSPR}}(n)-1}$, and
		\item $ k_n \sim \frac{2}{\Delta_{\text{rSPR}}(n)-1}$.
	\end{enumerate}
\end{conj}
Proving or disproving this conjecture would go a long way toward understanding the effects of relative distance on curvature.
However, we suspect that this will require a greater understanding of the distribution of pairwise distances between tree neighborhoods than is currently known.

@@

\begin{lem}
\label{lem:curvature_lower_bound}
	The curvature between adjacent trees with $n$ leaves is at least
	$$\frac{-n^2 + 2n + 1}{3.5n^2 - 15n + 16}.$$
\end{lem}
\begin{proof}
    In light of Corollary \ref{cor:paired_neighbors}, the optimal mass transport cost is maximized (and therefore curvature minimized) across adjacent trees $T$ and $S$ by a combination of two effects: trees that cannot be paired at distance $1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
		As we will show, these effects can be maximized simultaneously.
    To bound these effects, let $\mu$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
		Pairs of neighbors of adjacent trees are at most distance 3 apart, so $\mu$ of the mass is moved a distance at most 3.
		The remaining $1-\mu$ of the mass is moved between adjacent trees.
		Thus, we can bound the mass transport cost from above by $3\mu + 1(1-\mu) = 1 + 2\mu$.
	This gives a lower bound of $1 - (1 + 2\mu) / 1) = -2\mu$ on the curvature.

	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
	By Corollary~\ref{cor:paired_neighbors}, there are at least $\overlap := \degree{T} -2ka - 2(i-1)$ paired trees, bounding the former effect.
	We now construct a pair of trees that maximizes both effects.
	Let $S$ be the ladder tree with degree $3n^2 - 13n + 14$ and $T$ be the adjacent tree constructed by moving the lower $\floor{\frac{n}{2}}$ leaves of $S$ to the root.
	$T$ has degree at most $3.5n^2 -15n + 16$.
	There are thus $2ka + 2(i-1) = 2\left(\floor{\frac{n}{2}}\ceil{\frac{n-2}{2}} +(1-1)\right) \le \frac{1}{2}n^2 - n + \frac{1}{2}$ unpaired neighbors, the maximum possible.
	Moreover, as shown by Lemma~\ref{lem:degree_change} this pair of trees obtains the maximum (absolute and relative) degree change.
	Thus, the maximum $\mu$ is:
	$$\frac{\frac{1}{2}n^2 - n + \frac{1}{2}}{3.5n^2 - 15n + 16}.$$
	Note that any pair of trees with a smaller degree change will necessarily reduce the numerator by at least the same amount as the denominator, allowing us to use the degree of $T$ here rather than assuming the minimum degree of $S$.
	The claim follows from multiplying this value by $-2$.
\end{proof}

@@

Note that this bound is not tight (at least for small $n$) as it is rarely necessary to transport mass the maximum distance between unpaired trees.
We also note that the lower bounds in this section do not follow from the more general setting described in \citep{Jost2013-ce}.
However, the pair of trees used in the proof of Lemma~\ref{lem:curvature_lower_bound} will always have negative curvature, for all $n \ge 7$.
This follows by assuming that each of the unpaired trees is moved the minimum distance of one, that is the curvature for this pair of trees falls between $-\mu$ and $-2\mu$.

@@

	For the upper bound, we observe that $W_1'(T,S) \ge$
	\begin{align*}
		p &W_1(T,S) + (1-p) \, \dspr{T,S} - \frac{2p}{\max(\degree{T}, \degree{S})},
	\end{align*}
	as at most $p / \max(\degree{T},\degree{S})$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
	The upper bound then follows analogously to the lower bound.
	Moreover, no mass can remain at $T$ or $S$ when $\dspr{T,S} > 1$, in which case the curvatures are equal.
\end{proof}

@@

\section{Conclusion and future work}
In summary, we have gone beyond graph diameter and vertex degree to substantially advance the understanding of the phylogenetic rSPR graph.
We did so by developing the first theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
We found that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
Moreover, we proved that rSPR graph degree changes depend on the product of the size of the regrafted subtree with its change in depth.
This product is quadratic in the number of leaves in the trees, the same order as the degree.
As a result of this dependence, some areas of the rSPR graph are locally negatively curved but the graph tends toward flatness with respect to the majority of rSPR moves which move small subtrees.
Finally, we proved that the coarse and asymptotic definitions of Ricci-Ollivier curvature are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
